---
title: "p8105_hw2_co2616"
output: github_document
date: "2025-10-01"
---


loading the libraries to complete HW2

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
```

## Problem 1

tidying `pols` dataset

```{r}
pols_df = 
  read_csv("hw2_data/pols-month.csv") %>% 
  janitor::clean_names() %>% 
  separate(
    mon, into = c("year", "month", "day")) %>% 
  mutate(
    month = case_match(
    month,
    "01" ~ "january",
    "02" ~ "february",
    "03" ~ "march",
    "04" ~ "april",
    "05" ~ "may",
    "06" ~ "june",
    "07" ~ "july",
    "08" ~ "august",
    "09" ~ "september",
    "10" ~ "october",
    "11" ~ "november",
    "12" ~ "december"
  )) %>% 
  pivot_longer(
    cols = c(prez_gop, prez_dem),
    names_to = "president",
    values_to = "party",
    names_prefix = "prez_"
  ) %>% 
  select(-day)
```

tidying `snp` dataset

```{r}
snp_df = 
  read_csv("hw2_data/snp.csv") %>% 
  janitor::clean_names() %>% 
  mutate(date = mdy(date)) %>% 
  separate(
    date, into = c("year", "month", "day")) %>%
  mutate(
    month = case_match(
    month,
    "01" ~ "january",
    "02" ~ "february",
    "03" ~ "march",
    "04" ~ "april",
    "05" ~ "may",
    "06" ~ "june",
    "07" ~ "july",
    "08" ~ "august",
    "09" ~ "september",
    "10" ~ "october",
    "11" ~ "november",
    "12" ~ "december")) %>% 
  relocate("year", "month") %>% 
  select(-day)
```

tidying `unemployment` dataset

```{r}
unemploy_df = 
  read_csv("hw2_data/unemployment.csv",
  col_types = cols(
    Year = col_character()
  )) %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    cols = jan:dec,
    names_to = "month",
    values_to = "percentage",
  ) %>% 
 mutate(
    month = case_match(
    month,
    "jan" ~ "january",
    "feb" ~ "february",
    "mar" ~ "march",
    "apr" ~ "april",
    "may" ~ "may",
    "jun" ~ "june",
    "jul" ~ "july",
    "aug" ~ "august",
    "sep" ~ "september",
    "oct" ~ "october",
    "nov" ~ "november",
    "dec" ~ "december")) 
```

joining/merging the datasets

```{r}
first_df = 
  left_join(snp_df, pols_df, by = c("year", "month")) %>% 
  left_join(unemploy_df, by = c("year", "month"))
```

The `pols` dataset contained 822 observations of 9 variables related to the number of national politicians who are democratic or republican at any given time. The `snp` dataset contained 787 observations of 2 variables related to Standard & Poor’s stock market index (S&P). The The `unemployment` dataset contained 68 observations of 13 variables related to percentage of unemployment. After merging all three datasets together the resulting dataset, `first_df`, contained 1,345 observations and 12 variables. The key variables were `year` and `month` and years ranged from 1969 to 2068.


## Problem 2

tidying Mr.Trash Wheel dataset

```{r}
mr_trash_df =
  read_excel("hw2_data/202409 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N653") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(sports_balls = as.integer(sports_balls)) %>% 
  mutate(year = as.integer(year)) %>% 
  mutate(trash = "mr_wheel")
```

tidying Professor Trash Wheel dataset

```{r}
prof_trash_df =
  read_excel("hw2_data/202409 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", skip = 1) %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(trash = "professor")
```

tidying Gwynnda dataset

```{r}
gwy_trash_df =
  read_excel("hw2_data/202409 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel", skip = 1) %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(trash = "gwynnda")

gwy_trash_df_filter =
  gwy_trash_df %>% 
  filter(month == "June", year == "2022")
```

combining all three datasets

```{r}
trash_df = 
  bind_rows(mr_trash_df, prof_trash_df, gwy_trash_df) %>% 
  janitor::clean_names()
```

The `mr_trash_df` dataset contained 651 observations of 15 variables, the `prof_trash_df` dataset contained 119 observations of 14 variables, and the `gwy_trash_df` dataset contained 263 observations of 13 varaibles. After combining all three datasets together the resulting dataset, `trash_df`, contained 1,033 observations and 15 variables. Example of key variables were `dumpster`, `month`, `year`, and `plastic_bottles`. The total weight of trash collected by Professor Trash Wheel is `r sum(pull(prof_trash_df, weight_tons), na.rm = TRUE)`. The total number of cigarette butts collected by Gwynnda in June of 2022 is `r sum(pull(gwy_trash_df_filter, cigarette_butts))`



